Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2017-03-20 paper/ms.tex
Some intro to section 5; minor edits

To scale GP models to larger datasets, \citet{Rybicki:1995} presented a method
of computing the first term in \eq{gp-likelihood} in $\mathcal{O}(N)$
operations when the covariance function is given by
\begin{eqnarray}\eqlabel{kernel-simple}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})
\end{eqnarray}
where $\{{\sigma_n}^2\}_{n=1}^N$ are the measurement uncertainties,
$\delta_{nm}$ is the Kronecker delta, and $\bvec{\alpha} = (a,\,c)$.
The intuition behind this method is that, for this choice of $k_\bvec{\alpha}$,
the inverse of $K_\bvec{\alpha}$ is tridiagonal and can it can be computed
with a small number of operations for each data point.
Subsequently, \citet{Ambikasaran:2015} generalized this method to arbitrary
mixtures of exponentials
\begin{eqnarray}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J a_j\,\exp(-c_j\,\tau_{nm})\quad.
\end{eqnarray}
In this case, the inverse will be dense but \eq{gp-likelihood} can still be
evaluated in $\mathcal{O}(N)$ operations where $J$ is the number of components
in the mixture and $N$ is still the number of data points.

@@

In the case of the full \celerite\ covariance function
(\eqalt{celerite-kernel}), we introduce the following auxiliary variables in
analogy to the $u_n$ and $g_n$ that we introduced in the previous section
\begin{eqnarray}\eqlabel{full-system-first}
\phi_{n,j} &=& e^{-c_j\,\tau_{n+1,n}}\,\cos\left(d_j\,\tau_{n+1,n}\right)\\
\psi_{n,j} &=& -e^{-c_j\,\tau_{n+1,n}}\,\sin\left(d_j\,\tau_{n+1,n}\right)\\
g_{n,j} &=& \phi_{n,j}\,g_{n-1,j} + \phi_{n,j}\,z_n + \psi_{n,j}\,h_{n-1,j}\\
h_{n,j} &=& \phi_{n,j}\,h_{n-1,j} - \psi_{n,j}\,z_n - \psi_{n,j}\,g_{n-1,j}\\
u_{n,j} &=& \phi_{n+1,j}\,u_{n+1,j} + a_j\,z_{n+1} + \psi_{n+1,j}\,v_{n+1,j}\\
v_{n,j} &=& \phi_{n+1,j}\,v_{n+1,j} - b_j\,z_{n+1} - \psi_{n+1,j}\,u_{n+1,j}
\end{eqnarray}
with the boundary conditions
\begin{eqnarray}
    g_{0,j} = 0 \quad, \quad
    h_{0,j} = 0 \quad, \quad
    u_{N,j} = 0 \quad, \quad\mathrm{and}\quad
    v_{N,j} = 0
\end{eqnarray}
for all $j$.
Using these variables and some algebra, we find that the following expression
\begin{eqnarray}\eqlabel{full-system-last}
\sum_{j=1}^J \left[a_j\,g_{n,j}+b_j\,h_{n,j}\right]
+ \left[{\sigma_n}^2+\sum_{j=1}^J a_j\right]
+ \sum_{j=1}^J \left[\phi_{n,j}\,u_{n,j}+\psi_{n,j}\,v_{n,j}\right]
    &=& r_{\bvec{\theta},n}
\end{eqnarray}
is equivalent to the target matrix equation
\begin{eqnarray}
K\,\bvec{z} &=& \bvec{r}_\bvec{\theta}
\end{eqnarray}
if $r_{\bvec{\theta},n}$ is the $n$-th element of the residual vector
$\bvec{r}_\bvec{\theta}$ defined in \sect{gps}.
\eq{full-system-first} through \eq{full-system-last} define a banded matrix
equation in the ``extended'' space and, as before, this can be used to solve
for $K^{-1}\,\bvec{r}_\bvec{\theta}$ and $\det K$ in $\mathcal{O}(N)$
operations.
\Figure{matrix} shows a pictorial representation of the sparsity pattern of
the extended matrix $K_\mathrm{ext}$.
Given that definition of $K_\mathrm{ext}$, the corresponding extended vectors
$\bvec{z}_\mathrm{ext}$ and $\bvec{r}_\mathrm{ext}$ are defined schematically
as
\begin{eqnarray}
{\bvec{z}_\mathrm{ext}} ^\T =
\left(\begin{array}{cccccccccc}
    z_1 & u_{1,j} & v_{1,j} & g_{1,j} & h_{1,j} & z_2 & u_{2,j} & \cdots &
    h_{N-1,j} & z_N
\end{array}\right)
\end{eqnarray}
and
\begin{eqnarray}
{\bvec{r}_\mathrm{ext}} ^\T =
\left(\begin{array}{cccccccccc}
    r_{\bvec{\theta},1} & 0 & 0 & 0 & 0 & r_{\bvec{\theta},2} &
    0 & \cdots & 0 & r_{\bvec{\theta},2}
\end{array}\right) \quad.
\end{eqnarray}
After constructing the extended matrix (using a compact storage format), the
extended matrix can be factorized using a LU-decomposition\footnote{Even
though $K_\mathrm{ext}$ is symmetric, it is not positive definite so a
Cholesky solver cannot be used for increased efficiency.} routine optimized
for band or sparse matrices.
This decomposition can then be used to compute the determinant of $K$, solve
$K^{-1}\,\bvec{r}_\bvec{\theta}$, and subsequently calculate the marginalized
likelihood in \eq{gp-likelihood}.

@@

The extended system defined in the previous section is sparse with fewer than
a few percent non-zero entries and band structure.
In this \sectionname, we empirically investigate the performance and scaling
of three different algorithms for solving this extended system:

@@

\dfmtodo{Add more intro to this section.}

A special case of the \celerite\ model of great physical interest is a
stochastically-driven damped simple harmonic oscillator.
The differential equation for this system is
\begin{equation}
    \left[\frac{\dd^2}{\dd t^2} + \frac{\omega_0}{Q}\,\frac{\dd}{\dd t}
    + \omega_0^2\right]\, y(t) = \epsilon(t)
\end{equation}
where $\omega_0$ is the frequency of the undamped oscillator, $Q$ is the
quality factor of the oscillator, and $\epsilon(t)$ is a stochastic driving
force.
If $\epsilon(t)$ is white noise, the PSD of this process is given by
\citep[for example][]{Anderson:1990}
\begin{equation}\eqlabel{sho-psd}
S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0\,\omega_0^4}
    {(\omega^2-\omega_0^2)^2 + \omega_0^2\omega^2/Q^2}
\end{equation}
where $S_0$ is proportional to the power at $\omega = \omega_0$, $S(\omega_0)
= \sqrt{2/\pi}\,S_0\,Q^2$.
The power spectrum in \eq{sho-psd} matches \eq{celerite-psd} if
\begin{eqnarray}\eqlabel{sho-complex}
a_j &=& S_0\,\omega_0\,Q \\
b_j &=& \frac{S_0\,\omega_0\,Q}{\sqrt{4\,Q^2-1}} \\
c_j &=& \frac{\omega_0}{2\,Q}\\
d_j &=& \frac{\omega_0}{2\,Q} \sqrt{4\,Q^2-1} \quad,
\end{eqnarray}
for $Q \ge \frac{1}{2}$.
For $0 < Q \le \frac{1}{2}$, \eq{sho-psd} can be captured by a pair of \celerite\
terms with parameters
\begin{eqnarray}\eqlabel{sho-real}
a_{j\pm} &=& \frac{1}{2}\,S_0\,\omega_0\,Q\,\left[ 1 \pm
        \frac{1}{\sqrt{1-4\,Q^2}}\right] \\
b_{j\pm} &=& 0 \nonumber\\
    c_{j\pm} &=& \frac{\omega_0}{2\,Q}\,\left[1 \mp \sqrt{1-4\,Q^2}\right]
    \nonumber\\
d_{j\pm} &=& 0 \quad. \nonumber
\end{eqnarray}

>>>>

To scale GP models to larger datasets, \citet{Rybicki:1995} presented a method
of computing the first term in \eq{gp-likelihood} in $\mathcal{O}(N)$
operations when the covariance function is given by
\begin{eqnarray}\eqlabel{kernel-simple}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})
\end{eqnarray}
where $\{{\sigma_n}^2\}_{n=1}^N$ are the measurement uncertainties,
$\delta_{nm}$ is the Kronecker delta, and $\bvec{\alpha} = (a,\,c)$.
The intuition behind this method is that, for this choice of $k_\bvec{\alpha}$,
the inverse of $K_\bvec{\alpha}$ is tridiagonal and can be computed
with a small number of operations for each data point.
Subsequently, \citet{Ambikasaran:2015} generalized this method to arbitrary
mixtures of exponentials
\begin{eqnarray}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J a_j\,\exp(-c_j\,\tau_{nm})\quad.
\end{eqnarray}
In this case, the inverse will be dense but \eq{gp-likelihood} can still be
evaluated in $\mathcal{O}(N)$ operations where $J$ is the number of components
in the mixture and $N$ is still the number of data points.

@@

In the case of the full \celerite\ covariance function
(\eqalt{celerite-kernel}), we introduce the following auxiliary variables in
analogy to the $u_n$ and $g_n$ that we introduced in the previous section
\begin{eqnarray}\eqlabel{full-system-first}
\phi_{n,j} &=& e^{-c_j\,\tau_{n+1,n}}\,\cos\left(d_j\,\tau_{n+1,n}\right)\\
\psi_{n,j} &=& -e^{-c_j\,\tau_{n+1,n}}\,\sin\left(d_j\,\tau_{n+1,n}\right)\\
g_{n,j} &=& \phi_{n,j}\,g_{n-1,j} + \phi_{n,j}\,z_n + \psi_{n,j}\,h_{n-1,j}\\
h_{n,j} &=& \phi_{n,j}\,h_{n-1,j} - \psi_{n,j}\,z_n - \psi_{n,j}\,g_{n-1,j}\\
u_{n,j} &=& \phi_{n+1,j}\,u_{n+1,j} + a_j\,z_{n+1} + \psi_{n+1,j}\,v_{n+1,j}\\
v_{n,j} &=& \phi_{n+1,j}\,v_{n+1,j} - b_j\,z_{n+1} - \psi_{n+1,j}\,u_{n+1,j}
\end{eqnarray}
with the boundary conditions
\begin{eqnarray}
    g_{0,j} = 0 \quad, \quad
    h_{0,j} = 0 \quad, \quad
    u_{N,j} = 0 \quad, \quad\mathrm{and}\quad
    v_{N,j} = 0
\end{eqnarray}
for all $j$.
Using these variables and some algebra, we find that the following expression
\begin{eqnarray}\eqlabel{full-system-last}
\sum_{j=1}^J \left[a_j\,g_{n,j}+b_j\,h_{n,j}\right]
+ \left[{\sigma_n}^2+\sum_{j=1}^J a_j\right]
+ \sum_{j=1}^J \left[\phi_{n,j}\,u_{n,j}+\psi_{n,j}\,v_{n,j}\right]
    &=& r_{\bvec{\theta},n}
\end{eqnarray}
is equivalent to the target matrix equation
\begin{eqnarray}
K\,\bvec{z} &=& \bvec{r}_\bvec{\theta}
\end{eqnarray}
if $r_{\bvec{\theta},n}$ is the $n$-th element of the residual vector
$\bvec{r}_\bvec{\theta}$ defined in \sect{gps}.
\eq{full-system-first} through \eq{full-system-last} define a banded matrix
equation in the ``extended'' space and, as before, this can be used to solve
for $K^{-1}\,\bvec{r}_\bvec{\theta}$ and $\det K$ in $\mathcal{O}(N)$
operations.
\Figure{matrix} shows a pictorial representation of the sparsity pattern of
the extended matrix $K_\mathrm{ext}$.
Given that definition of $K_\mathrm{ext}$, the corresponding extended vectors
$\bvec{z}_\mathrm{ext}$ and $\bvec{r}_\mathrm{ext}$ are defined schematically
as
\begin{eqnarray}
{\bvec{z}_\mathrm{ext}} ^\T =
\left(\begin{array}{cccccccccc}
    z_1 & u_{1,j} & v_{1,j} & g_{1,j} & h_{1,j} & z_2 & u_{2,j} & \cdots &
    h_{N-1,j} & z_N
\end{array}\right)
\end{eqnarray}
and
\begin{eqnarray}
{\bvec{r}_\mathrm{ext}} ^\T =
\left(\begin{array}{cccccccccc}
    r_{\bvec{\theta},1} & 0 & 0 & 0 & 0 & r_{\bvec{\theta},2} &
    0 & \cdots & 0 & r_{\bvec{\theta},N}
\end{array}\right) \quad.
\end{eqnarray}
After constructing the extended matrix (using a compact storage format), the
extended matrix can be factorized using a LU-decomposition\footnote{Even
though $K_\mathrm{ext}$ is symmetric, it is not positive definite so a
Cholesky solver cannot be used for increased efficiency.} routine optimized
for band or sparse matrices.
This decomposition can then be used to compute the determinant of $K$, solve
$K^{-1}\,\bvec{r}_\bvec{\theta}$, and subsequently calculate the marginalized
likelihood in \eq{gp-likelihood}.

@@

The extended system defined in the previous section is sparse with typically fewer than
a few percent non-zero entries and band structure.
In this \sectionname, we empirically investigate the performance and scaling
of three different algorithms for solving this extended system:

@@

We now turn to the application of the \celerite\ algorithm to astrophysical
variability.  Many astrophysical objects have a characteristic timescale
of variability which is determined by its physical structure. The variability
is excited by noisy physical processes, but also damped due to dissipation 
in the system.  As a concrete example, stellar oscillations are strong at 
resonant frequencies determined by the internal stellar structure,
which are excited and damped by convective turbulence.

Thus, a special case of the \celerite\ model of great physical interest is a
stochastically-driven damped simple harmonic oscillator.
The differential equation for this system is
\begin{equation}
    \left[\frac{\dd^2}{\dd t^2} + \frac{\omega_0}{Q}\,\frac{\dd}{\dd t}
    + \omega_0^2\right]\, y(t) = \epsilon(t)
\end{equation}
where $\omega_0$ is the frequency of the undamped oscillator, $Q$ is the
quality factor of the oscillator, and $\epsilon(t)$ is a stochastic driving
force.
If $\epsilon(t)$ is white noise, the PSD of this process is given by
\citep[for example][]{Anderson:1990}
\begin{equation}\eqlabel{sho-psd}
S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0\,\omega_0^4}
    {(\omega^2-\omega_0^2)^2 + \omega_0^2\omega^2/Q^2}
\end{equation}
where $S_0$ is proportional to the power at $\omega = \omega_0$, $S(\omega_0)
= \sqrt{2/\pi}\,S_0\,Q^2$.
The power spectrum in \eq{sho-psd} matches \eq{celerite-psd} if
\begin{eqnarray}\eqlabel{sho-complex}
a_j &=& S_0\,\omega_0\,Q \\
b_j &=& \frac{S_0\,\omega_0\,Q}{\sqrt{4\,Q^2-1}} \\
c_j &=& \frac{\omega_0}{2\,Q}\\
d_j &=& \frac{\omega_0}{2\,Q} \sqrt{4\,Q^2-1} \quad,
\end{eqnarray}
for $Q \ge \frac{1}{2}$.
For $0 < Q \le \frac{1}{2}$, \eq{sho-psd} can be captured by a pair of \celerite\
terms with parameters
\begin{eqnarray}\eqlabel{sho-real}
a_{j\pm} &=& \frac{1}{2}\,S_0\,\omega_0\,Q\,\left[ 1 \pm
        \frac{1}{\sqrt{1-4\,Q^2}}\right] \\
b_{j\pm} &=& 0 \nonumber\\
    c_{j\pm} &=& \frac{\omega_0}{2\,Q}\,\left[1 \mp \sqrt{1-4\,Q^2}\right]
    \nonumber\\
d_{j\pm} &=& 0 \quad. \nonumber
\end{eqnarray}
